Evaluating infectious disease forecasts \\ with allocation scoring rules

### 2024-02-21 alloscore_manuscript/alloscore-application-shorter.Rnw
finish discussion

Our example application of the AS refers to a hypothetical and unspecified resource need that is supposed to 
pair with our forecasted outcome (new hospital admissions) in a manner that might appear overly simplistic and 
artificially direct.
A concrete decision problem, which motivated our work on allocation scoring, is

allocation of ventilators during a respiratory viral pandemic
\citep{huang_stockpiling_2017}.
We used ventilators as a notional example for a resource allocation problem because it (a) has been identified as a
scarce resource during outbreaks \citep{huang_stockpiling_2017}, (b) could be allocated across different locations, with
logistical considerations depending on the geographical scales, and (c) would likely be correlated (although not
perfectly so) with hospitalization data. However, this example has limitations in that not everyone who is hospitalized
needs a ventilator, and new hospital admissions may not be the best stand in for a need that corresponds to a prevalence
of severe cases. Other possible real-world examples include the allocation of a limited stockpile of vaccinations
\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests \citep{du_optimal_2022,pasco_covid-19_2023}.

While
there may be limitations in this specific pairing of a resource and forecasted outcome, we emphasize that the important
result from this work is how evaluating forecasts using a fundamentally different kind of score can illuminate
previously hidden aspects of value in forecasts. In this setting, the feature of a forecast that emerges as being
important is that the probabilistic forecasts predict the relative magnitudes between locations accurately, even if the
absolute magnitudes are inaccurate.

@@

We address some technical and methodological points from the main text. We begin in section \ref{sec:proper} by defining proper scoring rules and showing that the allocation score is proper. In section \ref{sec:ex-shortage}, we formalize the concept of a \emph{shortage} of resources, give some key results about expected resource shortages under a distribution characterizing uncertainty about (future) levels of resource need, and illustrate how quantiles arise as the Bayes act in a decision making problem about the quantity of a resource to purchase. Section \ref{sec:bayes-quantiles} gives a justification for the result that the Bayes act for the allocation problem is given by a vector of quantiles of the forecast distributions in each location at a shared probability level, which was stated in section \ref{sec:methods.detailed.specific_allocation} of the main text. We describe the algorithm that we use to compute allocations given a forecast distribution in each location in section \ref{sec:numeric}. In section \ref{sec:distfromq}, we describe the methods for approximating a full distribution from a set of quantiles, implemented in the R package \verb`distfromq`, that we used to support computation of allocation scores from quantile forecasts that were submitted to the US COVID-19 Forecast Hub for the application in the main text. Finally, in section \ref{sec:propriety_of_parametric_approximation} we examine implications for propriety of an analysis that uses summaries of forecasts (such as predictive quantiles) rather than the full forecast distributions for computation of allocation scores.

@@

Our primary decision theoretical procedure, outlined in section \ref{sec:methods.detailed.decisiontheory} of the main text,
uses a decision problem with loss function $s(x,y)$ to define a scoring rule
\begin{align}
S(F,y) := s(x^F,y) \label{eqn:bayes_sr}
\end{align}
where $x^F := \argmin_{x} \Ex_F[s(x,Y)]$ is the Bayes act for $F$ with respect to $s$.
Such scoring rules, which we call \emph{Bayes scoring rules},
are proper by construction since
\begin{align}
\Ex_F [S(F, Y)] &= \Ex_F [ s(x^F, Y) ] \nonumber \\
 &= \mathrm{min}_{x} \Ex_F [ s(x, Y) ] \quad \text{ (by definition of $x^F$)} \\
 &\leq \Ex_F [ s(x^G, Y) ] \label{eqn:dt_proper_key} \\
 &= \Ex_F [ S(G, Y)]. \nonumber
\end{align}

@@

Here we study the form of the Bayes act for the allocation problem (AP) (equation \eqref{eqn:loss_fn} in section \ref{sec:methods.detailed.specific_allocation}) of the text:
\begin{align}
    \underset{0 \leq x}{\mathrm{minimize}}\,\, \mathbb{E}_{F} [s_A(x, Y)]= \sum_{i=1}^{N} L \cdot \mathbb{E}_{F_i}[(Y_i - x_i)_{+}]
     \text{ subject to }
     \, \sum_{i=1}^N x_i = K, \label{AP-long}
\end{align}
where the marginal forecasts $F_i$ for $i=1,\dots,N$ represent forecasts for $N$ distinct locations.
We show that the Bayes act $x^{F,K} = (x_1^{F,K},\ldots,x_N^{F,K})$ for a forecast $F$ and resource constraint level $K$
is a vector of quantiles of the marginal forecast distributions $F_i$ at a single probability level
$\tau^{F,K}$, that is, $x_i^{F,K} = q_{F_i,\tau^{F,K}}$. An immediate consequence used in the examples in Section \ref
{sec:methods.overview} in the main text is that if $F_i = \mathrm{Exp}(1/\sigma_i)$ for all $i$, then the Bayes act is
proportional to $(\sigma_1,\ldots,\sigma_N)$, since $q_{\mathrm{Exp}(1/\sigma),\tau} = -\sigma \log(1-\tau)$.

@@

We have implemented an iterative bisection method that makes use of the fact that $\sum_{i=1}^N q_{\tau,F_i}$
is an increasing function of $\tau$.
The algorithm begins with an initial search interval $[\tau_{L,1}, \tau_{U,1}]$ (such as $[0,\max_{i}F_i(K)]$) that clearly contains
the solution $\tau^{\star}$.
% This can be confirmed by verifying that the resource allocations corresponding to the lower endpoint of the search interval do not exceed the resource constraint ($\sum_{i=1}^N q_{\tau_L,F_i} \leq K$) and the resource allocations corresponding to the upper endpoint match or exceed the available resources ($\sum_{i=1}^N q_{\tau_U,F_i} \geq K$).
At each step $j$ of the algorithm, we evaluate the total allocation $\sum_{i=1}^N q_{\tau_{M,j},F_i}$ at the midpoint of the search interval,
$\tau_{M,j} = \frac{1}{2}(\tau_{L,j} + \tau_{U,j})$ and continue the search on the narrowed sub-interval
\begin{align}
[\tau_{L,j+1},\tau_{U,j+1}] =
\begin{cases}
[\tau_{L,j}, \tau_{M,j}] & \text{if } \sum_{i=1}^N q_{\tau_{M,j},F_i} \geq K \\
[\tau_{M,j}, \tau_{U,j}] & \text{if } \sum_{i=1}^N q_{\tau_{M,j},F_i} < K.
\end{cases} \nonumber
\end{align}
% We then narrow the search to the sub-interval that contains the solution:
% \begin{itemize}
%   \item If $\sum_{i=1}^N q_{\tau_{M},F_i} \geq K$, we continue the search on the interval $[\tau_L, \tau_M]$.
%   \item If $\sum_{i=1}^N q_{\tau_M, F_i} \leq K$, we continue the search on the interval $[\tau_M, \tau_U]$.
% \end{itemize}
This search continues until $\tau_{U,j+1} < (1+\varepsilon)\tau_{L,j+1}$ for a suitably small $\varepsilon>0$.
We have implemented this procedure along with the resulting score computations in the
R package \verb`alloscore` \citep{gerding-alloscore} which provided all allocation score values used in the analysis of section \ref{sec:application}
in the main text.

@@

We find it helpful now to formally distinguish between two decision making problems. The first is the public health decision maker's allocation problem where the task is to select an allocation $x$, with the allocation loss $s_A(x, y) = \sum_{i=1}^N L \cdot \max(0, y_i - x_i)$ as described in section \ref{sec:methods.detailed.specific_allocation} of the main text.
The second is the forecaster's reporting problem where the task is to select parameter values $\theta$ to report. The forecaster's loss is given by
\begin{align}
s_R(\theta, y) = s_A(x^{G_\theta}, y), \label{eqn:forecaster_theta_loss}
\end{align}
where $x^{G_\theta}$ is the Bayes act for the allocation problem under the distribution $G_\theta$. In words, the loss associated with reporting $\theta$ is equal to the loss associated with taking the Bayes allocation corresponding to the distribution $G_\theta$.

>>>>

Our example application of the AS refers to a hypothetical and unspecified resource need that is assumed to 
pair with our forecasted outcome (new hospital admissions) in a manner that might appear overly simplistic and 
artificially direct.  We are optimistic, however, that by augmenting this scheme with a probabilistic measurement
model for resource need and more detailed domain knowledge, it could be meaningfully aligned with a 
more concrete decision problem such as the allocation of ventilators during a respiratory viral pandemic.  This problem 
was explored from a stochastic optimization perspective (not from a forecast evaluation perspective) in 
\cite{huang_stockpiling_2017} and provided an initial motivation for this project.
We see it as promising focus for further development of the AS framework since it (a) has been identified as a prime
manifestation of scarcity during respiratory outbreaks (b) tansparently involves
logistical considerations depending on geographical and population scales, and (c) would likely imply a central role for 
hospitalization data. But again, additional modeling and data synthesis would be required since not everyone who is hospitalized
needs a ventilator, and new hospital admissions will not fully describe for a need that corresponds to a prevalence
of severe cases. Other possible real-world examples include the allocation of a limited stockpile of vaccinations
\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests \citep{du_optimal_2022,pasco_covid-19_2023}.

@@

We address some technical and methodological points from the main text. We begin in section \ref{sec:proper} by defining proper scoring rules and showing that the allocation score is proper. In section \ref{sec:ex-shortage}, we formalize the concept of a \emph{shortage} of resources, give some key results about expected resource shortages under a distribution characterizing uncertainty about (future) levels of resource need, and illustrate how quantiles arise as the Bayes act in a decision making problem about the quantity of a resource to purchase. Section \ref{sec:bayes-quantiles} gives a justification for the result that the Bayes act for the allocation problem is given by a vector of quantiles of the forecast distributions in each location at a shared probability level, which was stated in section 
\ref{sec:methods.detailed.specific_allocation}. We describe the algorithm that we use to compute allocations given a forecast distribution in each location in section \ref{sec:numeric}. In section \ref{sec:distfromq}, we describe the methods for approximating a full distribution from a set of quantiles, implemented in the R package \verb`distfromq`, that we used to support computation of allocation scores from quantile forecasts that were submitted to the US COVID-19 Forecast Hub for the application in section \ref{sec:application}. Finally, in section 
\ref{sec:propriety_of_parametric_approximation} we examine implications for propriety of an analysis that uses summaries of forecasts (such as predictive quantiles) rather than the full forecast distributions for computation of allocation scores.

@@

Our primary decision theoretical procedure, outlined in section \ref{sec:methods.detailed.decisiontheory},
uses a decision problem with loss function $s(x,y)$ to define a scoring rule
\begin{align}
S(F,y) := s(x^F,y) \label{eqn:bayes_sr}
\end{align}
where $x^F := \argmin_{x} \Ex_F[s(x,Y)]$ is the Bayes act for $F$ with respect to $s$.
Such scoring rules, which we call \emph{Bayes scoring rules},
are proper by construction since
\begin{align}
\Ex_F [S(F, Y)] &= \Ex_F [ s(x^F, Y) ] \nonumber \\
 &= \mathrm{min}_{x} \Ex_F [ s(x, Y) ] \quad \text{ (by definition of $x^F$)} \\
 &\leq \Ex_F [ s(x^G, Y) ] \label{eqn:dt_proper_key} \\
 &= \Ex_F [ S(G, Y)]. \nonumber
\end{align}

@@

Here we study the form of the Bayes act for the allocation problem (AP) (equation \eqref{eqn:loss_fn} in section \ref{sec:methods.detailed.specific_allocation}) of the text:
\begin{align}
    \underset{0 \leq x}{\mathrm{minimize}}\,\, \mathbb{E}_{F} [s_A(x, Y)]= \sum_{i=1}^{N} L \cdot \mathbb{E}_{F_i}[(Y_i - x_i)_{+}]
     \text{ subject to }
     \, \sum_{i=1}^N x_i = K, \label{AP-long}
\end{align}
where the marginal forecasts $F_i$ for $i=1,\dots,N$ represent forecasts for $N$ distinct locations.
We show that the Bayes act $x^{F,K} = (x_1^{F,K},\ldots,x_N^{F,K})$ for a forecast $F$ and resource constraint level $K$
is a vector of quantiles of the marginal forecast distributions $F_i$ at a single probability level
$\tau^{F,K}$, that is, $x_i^{F,K} = q_{F_i,\tau^{F,K}}$. An immediate consequence used in the examples in Section \ref
{sec:methods.overview} is that if $F_i = \mathrm{Exp}(1/\sigma_i)$ for all $i$, then the Bayes act is
proportional to $(\sigma_1,\ldots,\sigma_N)$, since $q_{\mathrm{Exp}(1/\sigma),\tau} = -\sigma \log(1-\tau)$.

@@

We have implemented an iterative bisection method that makes use of the fact that $\sum_{i=1}^N q_{\tau,F_i}$
is an increasing function of $\tau$.
The algorithm begins with an initial search interval $[\tau_{L,1}, \tau_{U,1}]$ (such as $[0,\max_{i}F_i(K)]$) that clearly contains
the solution $\tau^{\star}$.
% This can be confirmed by verifying that the resource allocations corresponding to the lower endpoint of the search interval do not exceed the resource constraint ($\sum_{i=1}^N q_{\tau_L,F_i} \leq K$) and the resource allocations corresponding to the upper endpoint match or exceed the available resources ($\sum_{i=1}^N q_{\tau_U,F_i} \geq K$).
At each step $j$ of the algorithm, we evaluate the total allocation $\sum_{i=1}^N q_{\tau_{M,j},F_i}$ at the midpoint of the search interval,
$\tau_{M,j} = \frac{1}{2}(\tau_{L,j} + \tau_{U,j})$ and continue the search on the narrowed sub-interval
\begin{align}
[\tau_{L,j+1},\tau_{U,j+1}] =
\begin{cases}
[\tau_{L,j}, \tau_{M,j}] & \text{if } \sum_{i=1}^N q_{\tau_{M,j},F_i} \geq K \\
[\tau_{M,j}, \tau_{U,j}] & \text{if } \sum_{i=1}^N q_{\tau_{M,j},F_i} < K.
\end{cases} \nonumber
\end{align}
% We then narrow the search to the sub-interval that contains the solution:
% \begin{itemize}
%   \item If $\sum_{i=1}^N q_{\tau_{M},F_i} \geq K$, we continue the search on the interval $[\tau_L, \tau_M]$.
%   \item If $\sum_{i=1}^N q_{\tau_M, F_i} \leq K$, we continue the search on the interval $[\tau_M, \tau_U]$.
% \end{itemize}
This search continues until $\tau_{U,j+1} < (1+\varepsilon)\tau_{L,j+1}$ for a suitably small $\varepsilon>0$.
We have implemented this procedure along with the resulting score computations in the
R package \verb`alloscore` \citep{gerding-alloscore} which provided all allocation score values used in the analysis of section \ref{sec:application}.

@@

We find it helpful now to formally distinguish between two decision making problems. The first is the public health decision maker's allocation problem where the task is to select an allocation $x$, with the allocation loss $s_A(x, y) = \sum_{i=1}^N L \cdot \max(0, y_i - x_i)$ as described in section \ref{sec:methods.detailed.specific_allocation}.
The second is the forecaster's reporting problem where the task is to select parameter values $\theta$ to report. The forecaster's loss is given by
\begin{align}
s_R(\theta, y) = s_A(x^{G_\theta}, y), \label{eqn:forecaster_theta_loss}
\end{align}
where $x^{G_\theta}$ is the Bayes act for the allocation problem under the distribution $G_\theta$. In words, the loss associated with reporting $\theta$ is equal to the loss associated with taking the Bayes allocation corresponding to the distribution $G_\theta$.
